---
title: "House Prices in New York"
author: "Abtahee Islam"
format: 
  revealjs:
    auto-slide: 20000
    embed-resources: true
    transition: slide       
    theme: simple           
    slide-number: true
---

## Introduction {.smaller}

- **Topic: Which property and neighborhood features best predict house prices in New York?**  
- **Motivation:** New York has one of the most expensive housing markets in the world. Understanding its pricing system helps buyers and investors make smarter decisions.  
- **Goal:** Identify the key variables that influence house prices and compare how their effects differ across several predictive models.
- **Why it matters:** Property traits are complex! They jointly affect metrics like affordability and urban development.

---

## Dataset Overview {.smaller}

- **Data source:** New York housing sales dataset (`housing-prices-ge19.txt`)  
- **Size:** 1734 observations, 17 variables  
- **Response variable:** Sale price of each property (USD)  
- **Predictor variables:**  
  - *Property features:* `Lot.Size`, `Living.Area`, `Bedrooms`, `Bathrooms`, `Fireplaces`, `Age`, `Rooms`  
  - *Neighborhood features:* `Pct.College` (college educated residents %), `Waterfront`, `Central.Air`, `Sewer.Type`, `Fuel.Type`, `Land.Value`  
- **Main question:** How much do location and property attributes explain prices?  

```{r}
#| label: setup-vis
#| echo: false
#| include: false

library(data.table)   
library(tidyverse)
library(gt)
library(scales)
library(patchwork)
library(readr)
library(ggplot2)
library(randomForest)
library(dplyr)
```

```{r}

data <- read_delim("housing-prices-ge19.txt", delim = "\t")

```
---

## Initial Observations

- **Median values of predictor variables:**  
  - *Price* ≈ \$189,700
  - *Lot size* ≈ 0.37 acres  
  - *Area* ≈ 1632 square feet
  - *Number of bedrooms* ≈ 3
  - *Number of bathrooms* ≈ 2  

```{r}
#| label: price-dist-and-medians
#| fig-cap: "Median summary"
#| fig-width: 8
#| fig-height: 4

data |>
summarise(
median_price = median(Price, na.rm = TRUE),
median_lot   = median(Lot.Size, na.rm = TRUE),
median_area  = median(Living.Area, na.rm = TRUE),
median_beds  = median(Bedrooms, na.rm = TRUE),
median_baths = median(Bathrooms, na.rm = TRUE)
)

```

---

## Distribution of House Prices 
:::{.columns}

::: {.column width="50%"}
```{r, fig.width=18, fig.height=18}
library(ggplot2)
library(scales)
library(patchwork)

#Histogram
ggplot(data, aes(x = Price)) +
  geom_histogram(binwidth = 10000, fill = "pink", color = "white") +
  labs(title = "Histogram of house prices", x = "Price (USD)", y = "Count") +
  scale_x_continuous(labels = scales::label_comma()) +
  theme(plot.title = element_text(size = 50),
        axis.text=element_text(size=18),
        axis.title=element_text(size=40))+
  geom_vline(aes(xintercept = median(Price, na.rm = TRUE)), color = "red", linetype = "dashed", size = 2)

```
:::

::: {.column width="50%"}
- Prices are **right-skewed**; most homes are in lower ranges.
- **Several high-value outliers** above $500,000.
- **Log transformation** recommended to stabilize variance.
- **Outlier handling** improves model robustness.
:::

:::
---

## Exploring Key Predictors of House Prices
```{r}
data_transformed <- data |>
filter(!is.na(Price), Price > 0) |>
mutate(
Price_1K = Price / 1000,
Waterfront = as.factor(Waterfront),
Central.Air = as.factor(Central.Air),
New.Construct = as.factor(New.Construct),
log_price = log(Price)
)

p1 <- ggplot(data_transformed, aes(x = Living.Area, y = Price)) +
geom_point(alpha = 0.5, color = "#1f78b4") +
geom_smooth(method = "lm", color = "#e31a1c", linewidth = 1) +
scale_y_continuous(labels = label_dollar()) +
labs(title = "Larger Living Areas Strongly Increase Price",
x = "Living Area (sq ft)",
y = "Price ($)") +
theme_minimal(base_size = 11) +
theme(plot.title = element_text(face = "bold", size = 13))


p2 <- ggplot(data_transformed, aes(x = factor(Bathrooms), y = Price)) +
geom_boxplot(fill = "#a6cee3", color = "#1f78b4") +
scale_y_continuous(labels = label_dollar()) +
labs(title = "Homes with More Bathrooms Cost More",
x = "Number of Bathrooms",
y = "Price ($)") +
theme_minimal(base_size = 11) +
theme(plot.title = element_text(face = "bold", size = 13))

p3 <- ggplot(data_transformed, aes(x = Waterfront, y = Price, fill = Waterfront)) +
geom_boxplot() +
scale_fill_manual(values = c("#b2df8a", "#33a02c")) +
scale_y_continuous(labels = label_dollar()) +
labs(title = "Waterfront Properties Command Higher Prices",
x = "Waterfront",
y = "Price ($)") +
theme_minimal(base_size = 11) +
theme(legend.position = "none",
plot.title = element_text(face = "bold", size = 13))

(p1 / (p2 | p3)) + plot_annotation(tag_levels = 'A')
```
*In our exploratory data analysis (EDA), we examined relationships between key predictors and house prices.*
---

## Data Cleaning & Transformation
```{r}
data_transformed = data |>
  mutate(Price_1K = Price / 1000) 

data_transformed$Waterfront <- as.factor(data_transformed$Waterfront)
data_transformed$Central.Air <- as.factor(data_transformed$Central.Air)
data_transformed$New.Construct <- as.factor(data_transformed$New.Construct)
```

<div style="font-size:30px;">
- Removed missing or invalid entries.
- Encoded categorical variables: Central.Air, Waterfront, New.Construct.
- Applied **log transformation** to stabilize variance.
- Data is now clean and ready for **multiple regression modelling**.
</div>

```{r}
data_transformed <- data |>
filter(!is.na(Price), Price > 0) |>
mutate(
Price_1K = Price / 1000,
Waterfront = as.factor(Waterfront),
Central.Air = as.factor(Central.Air),
New.Construct = as.factor(New.Construct),
log_price = log(Price)
)
```

```{r,  fig.width=10, fig.height=6}
ggplot(data_transformed, aes(x = log_price)) +
geom_histogram(binwidth = 0.1, fill = "steelblue", color = "white") +
labs(title = "Log-Transformed Price Distribution", x = "log(Price)", y = "Count") +
theme_minimal(base_size = 12) +
theme(
  plot.title = element_text(size = 20, face = "bold"),
  axis.title = element_text(size=14, face = "bold")
  )
```
---

## Correlation Matrix & Predictor Strength
```{r}

library(GGally)

vars_to_use <- c("Living.Area", "Land.Value", "Waterfront",
                 "Pct.College", "Bathrooms", "Age", 
                 "Bedrooms", "Lot.Size")

# Convert Waterfront to factor or numeric as needed
data_corr <- data_transformed %>% 
  dplyr::select(all_of(vars_to_use)) %>%
  dplyr::mutate(Waterfront = as.factor(Waterfront))

ggpairs(data_corr, title = "Scatterplot & Correlation Matrix") + 
  theme_bw()+
   theme(
    axis.text.x = element_text(size = 4),  
    axis.text.y = element_text(size = 4),  
    plot.title = element_text(size = 16),   
    strip.text = element_text(size = 6)    
   )
```

---

### List R² of the variables(simple linear regression)

```{r, echo=FALSE}
#Computing R-squared value

r2_living_area = summary(lm(Price ~ Living.Area, data = data))$r.squared
r2_land_value = summary(lm(Price ~ Land.Value, data = data))$r.squared
r2_waterfront = summary(lm(Price ~ Waterfront, data = data))$r.squared
r2_pct_college = summary(lm(Price ~ Pct.College, data = data))$r.squared
r2_bathrooms = summary(lm(Price ~ Bathrooms, data = data))$r.squared
r2_age = summary(lm(Price ~ Age, data = data))$r.squared
r2_bedrooms = summary(lm(Price ~ Bedrooms, data = data))$r.squared
r2_lot_size = summary(lm(Price ~ Lot.Size, data = data))$r.squared

r2_results = data.frame(
  Variable = c("Living.Area", "Land.Value", "Waterfront", "Pct.College", 
               "Bathrooms", "Age", "Bedrooms", "Lot.Size"),
  R_Squared = c(r2_living_area, r2_land_value, r2_waterfront, r2_pct_college,
                r2_bathrooms, r2_age, r2_bedrooms, r2_lot_size)
)

r2_results = r2_results[order(-r2_results$R_Squared), ]
r2_results
```
###### Key Findings:
<div style="font-size:36px;">
- Living Area has the highest R² (0.51), explaining 51% of price variance

- Land Value is second strongest predictor (R² = 0.34)

- We chose statistical variables for multiple linear regression model.
</div>

---

### Multiple Linear Regression Testing:

```{r}
df_data = data_transformed[, c("Price_1K", "Living.Area", "Land.Value", "Age", "Pct.College")]
cor_data = cor(df_data)


multi_model = lm(Price_1K ~ Living.Area + Land.Value + Age + Pct.College, data = data_transformed)
multi_summary_model = summary(multi_model)
multi_summary_model


```
<div style="font-size:32px;">
- Median is -5.16, which is very close to 0.
- Housing prices is strongly related with living area(0.71) and land value (0.58) and has a weak negative correlation with age(-0.19).
- The variable can explain 61% of changes of the housing prices.
</div>

---

## Diagnostic Plots and Assumption check

::::: columns
::: {.column width="50%"}
```{r, fig.height=6}
resid_model = resid(multi_model)
fitted_model = fitted(multi_model)

plot(fitted_model, resid_model, xlab = "Fitted Values", ylab = "Residuals", main = "Residuals vs Fitted values")
abline(h = 0, col = "red")

qqnorm(resid_model)
qqline(resid_model, col = "red")

```
:::

::: {.column width="50%"}
- Residual Plot: Distribution is relatively constant, the variance homogeneity and linear relationship assumptions are mostly satisfied.

- Q-Q plot: The middle part is normal, but there is a significant deviation in the right tail.
:::
:::::
---

## Random Forest Model

::::: columns
::: {.column width="60%"}
```{r}
set.seed(123)
rf_model = randomForest(Price ~ Bathrooms, data = data, ntree = 100)

bathroom_levels = data.frame(Bathrooms = c(1, 1.5, 2, 2.5, 3, 3.5, 4))
predictions = predict(rf_model, newdata = bathroom_levels)

result_df = data.frame(
  Bathrooms = bathroom_levels$Bathrooms,
  Predicted_Price = round(predictions)
)

result_df$Price_Growth = c(0, round(diff(result_df$Predicted_Price) / result_df$Predicted_Price[1:(nrow(result_df)-1)] * 100, 1))
result_df$Price_Growth[1] = NA 

rf_r2 = cor(data$Price, predict(rf_model, data))^2
lm_r2 = summary(lm(Price ~ Bathrooms, data = data))$r.squared

print(rf_model)

print(result_df)


```
:::

::: {.column width="40%"}
<div style="font-size:30px;">
- Non-linear approach captures relationship between bathroom and housing prices.
- Prices increase with bathroom count but not linearly.
- The price increase is most significant from 1.5 to 2.5 bathrooms, with diminishing returns starting with 3 bathrooms, indicating diminishing marginal returns.
</div>
:::
:::::

---

## Model Comparison

<div style="font-size:32px;">
- We compared a **multiple linear regression** (Price ~ Living Area + Land Value + Age + Pct.College) and a **Random Forest** model.

- Both models capture that larger homes and higher land values raise prices.

- The **Linear Regression** achieved a higher R², meaning it explained **more variance** and captured the main **linear trends** in price.

  - Linear Model R²: `r round(summary(multi_model)$r.squared, 4)`  
  - Random Forest R²: `r round(rf_r2, 4)`

- Overall, the linear model provided a strong, interpretable fit, while the Random Forest highlights how some non-linear effects may still exist in housing prices.
</div>

---

## Ranking Key Predictors
::::: {.columns}

::: {.column width="40%"}
```{r}
df <- data_transformed
y <- "Price"
predictors <- c("Living.Area","Land.Value", "Age","Pct.College")

form <- as.formula(paste(y, "~", paste(predictors, collapse = " + ")))
mod  <- lm(form, data = df)

coef_table <- summary(mod)$coefficients
t_stats <- abs(coef_table[-1, 3])

rank_df <- data.frame(
  Predictor   = rownames(coef_table)[-1],
  t_statistic = round(t_stats, 3)
)
rank_df <- rank_df[order(-rank_df$t_statistic), ]
rank_df <- head(rank_df, 4)

pretty_names <- c("Land.Value"="Land Value","Living.Area"="Living Area")
rank_df$Predictor <- ifelse(rank_df$Predictor %in% names(pretty_names),
                            pretty_names[rank_df$Predictor], rank_df$Predictor)

knitr::kable(rank_df, row.names = FALSE,
             caption = "Predictors ranked by |t-statistic|") |>
  kableExtra::kable_styling(font_size = 36)
```
:::

::: {.column width="60%"}
<div style="font-size:32px;">
- Predictors are ranked by the absolute value of the test statistics from the multiple linear regression model.

- Higher |t| values indicate stronger evidence that a variable affects price.

- **Land Value** and **Living Area** are the strongest predictors which show that property size and land worth are the main drivers of New York housing prices.
</div>
:::

:::::

---

## Predicted vs Actual Prices

::::: {.columns}
::: {.column width="60%"}

```{r}
data_transformed$Predicted_Price <- predict(multi_model, data_transformed)

price_comp <- data_transformed |>
  mutate(Price_Range = cut(Price_1K, breaks = c(0, 100, 200, 300, 400, 500, 1000), 
                           labels = c("<100", "100–200", "200–300", "300–400", "400–500", "500+"))) |>
  group_by(Price_Range) |>
  summarise(
    Actual = mean(Price_1K, na.rm = TRUE),
    Predicted = mean(Predicted_Price, na.rm = TRUE)
  ) |>
  pivot_longer(cols = c("Actual", "Predicted"), names_to = "Type", values_to = "Price")


ggplot(price_comp, aes(x = Price_Range, y = Price, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Predicted vs Actual Prices by Price Range",
    x = "Actual Price Range (in $1,000s)",
    y = "Average Price (in $1,000s)"
  ) +
  scale_fill_manual(values = c("Actual" = "red", "Predicted" = "lightblue")) +
  theme_minimal()
```
:::
::: {.column width="40%"}
<div style="font-size:26px;">

- The bar plot compares average predicted prices from the multiple linear regression model against actual house prices across different price ranges.

- The model fits mid-range properties (200–400K) quite well, showing similar predicted and actual prices.

- For lower-priced homes, predictions are slightly overestimated, while high-end homes (500K+) are underpredicted. 

</div>
:::
:::::

---

## Model Limitations
<div style= "font-size:32px;">
- The **Multiple Linear Regression model** assumes linear relationship between predictors and prices which does not hold in real-world markets as prices usually behave non-linearly.

<br>

- The **Random Forest model** captures non-linear relationships but it lacks transparency, making it difficult to interpret how each variable affects prices.

<br>

- Both models only used selected predictors, so not including other variables may reduce accuracy and leave out important factors influencing price.
</div>

---

## Key Insights

<div style="font-size:26px;">

### What We Found
- **Living Area** dominates; **Land Value** is next strongest  
- Bathrooms / Waterfront / Central Air ↑ price  
- **Age** ↓ price; **Pct.College** modest positive  
- RF reveals **non-linear patterns** and **diminishing returns**

### What It Means
- Price drivers center on **home size** and **land worth**  
- **Neighborhood quality** adds value, but physical traits dominate  
- Linear captures main trends; RF uncovers **non-linearities**
</div>

---

## Practical Implications

<div style="font-size:26px;">

### For Buyers
- Use key attributes to **benchmark fair value**  
- Spot listings that are **over-/under-priced**  
- Value-add: increase usable space / modernize age-sensitive features

### For Developers
- Prioritize **land acquisition** & **usable living area**  
- Extra rooms/baths beyond needs → **diminishing returns**  
- Newer construction tends to price higher (mitigate “age” effect)

### For Investors & Policy
- Target **high-education, fair-priced** neighborhoods  
- Support **affordability** and **zoning** assessments
</div>

---

## Conclusion

<div style="font-size:26px;">

### Summary of Findings

- Both **MLR** and **RF** agree: **Land Value, Living Area, Pct.College** matter most  
- **Linear** = interpretable, significant relationships  
- **RF** = non-linear effects & interactions  
- Together they explain **>60%** of price variation (in our sample)

### Key Takeaways
- Core economics: **scale + land**  
- **Statistics + ML** → accuracy **and** interpretability  
- Results inform buying, development, and investment choices
</div>

---

### Future Work
- Add **geo-coordinates** for spatial effects  
- Enrich **income / crime / school quality** variables  
- Include **time trends**; cross-city comparisons












